---
title: "Reproducible Research - Course Project 2"
output:
  pdf_document: default
  html_document: default
---

# Health and Economic Consequences of Weather Events in United States

# Synopsis
The report addresses questions regarding the most hurtful weather type events in terms of deaths, injuries, property and crop damage

# Data Processing
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load needed packages
```{r, echo=FALSE}
library(lubridate)
library(ggplot2)
library(dplyr)
library(rmarkdown)
library(plyr)
library(zoo)
```

```{r, echo=FALSE}
setwd("/Users/iopetrid/Desktop/Coursera/Data Science/5_Reproducible Research/Course Project 2")

storm<-read.csv("./repdata-data-StormData.csv.bz2",header = TRUE, sep=",")
str(storm)
head(tbl_df(storm))
```
## Questions

### 1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

```{r}
# Subset the data using only the values needed
var2use<-c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
storm<-storm[var2use]
str(storm)
head(tbl_df(storm))

# For this question I am going to find which EVTYPE leads to more FATALITES and INJURIES and present the 10 most dominant ones
total_deaths<-aggregate(FATALITIES ~ EVTYPE, data=storm, FUN = sum)

total_injuries<-aggregate(INJURIES ~ EVTYPE, data=storm, FUN = sum)

total_deaths_sorted<- total_deaths %>% arrange(-total_deaths$FATALITIES)

total_injuries_sorted<- total_injuries %>% arrange(-total_injuries$INJURIES)

most_fatal_events<-head(total_deaths_sorted,10)
most_fatal_events

par(mfrow = c(1,1), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0), cex = 0.8)
barplot(most_fatal_events$FATALITIES, names.arg = most_fatal_events$EVTYPE,las = 3,col="wheat",main="10 most fatal events",ylab="# of Deaths")
dev.copy(png, "fatal-events.png", width = 480, height = 480)
dev.off()

most_injuries_events<-head(total_injuries_sorted,10)
most_injuries_events

par(mfrow = c(1,1), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0), cex = 0.8)
barplot(most_injuries_events$INJURIES, names.arg = most_injuries_events$EVTYPE,las = 3,col="grey", main="10 events causing the most injuries",ylab="# of Injuries")
dev.copy(png, "injuries-events.png", width = 480, height = 480)
dev.off()
```

### 2. Across the United States, which types of events have the greatest economic consequences?
```{r}
# For this question I am going to find which EVTYPE lead to more property and crop damage (PROPDMGTOTAL,CROPDMGTOTAL) and present the 10 most fatal ones

#first I need to create a total variable for property and crop damage
# PROPDMGTOTAL=PROPDMG+PROPDMGEXP
storm$PROPDMGEXP <- mapvalues(storm$PROPDMGEXP, from = c("K", "M","", "B", "m", "+", "0", "5", "6", "?", "4", "2", "3", "h", "7", "H", "-", "1", "8"), to = c(10^3, 10^6, 1, 10^9, 10^6, 0,1,10^5, 10^6, 0, 10^4, 10^2, 10^3, 10^2, 10^7, 10^2, 0, 10, 10^8))
storm$PROPDMGEXP <- as.numeric(as.character(storm$PROPDMGEXP))
storm$PROPDMGTOTAL <- (storm$PROPDMG * storm$PROPDMGEXP)/1000000000

prop_damage<-aggregate(PROPDMGTOTAL ~ EVTYPE, data=storm, FUN = sum)

#CROPDMGTOTAL=CROPDMG+CROPDMGEXP
storm$CROPDMGEXP <- mapvalues(storm$CROPDMGEXP, from = c("","M", "K", "m", "B", "?", "0", "k","2"), to = c(1,10^6, 10^3, 10^6, 10^9, 0, 1, 10^3, 10^2))
storm$CROPDMGEXP <- as.numeric(as.character(storm$CROPDMGEXP))
storm$CROPDMGTOTAL <- (storm$CROPDMG * storm$CROPDMGEXP)/1000000000

crop_damage<-aggregate(CROPDMGTOTAL ~ EVTYPE, data=storm, FUN = sum)

total_prop_damage<- prop_damage %>% arrange(-prop_damage$PROPDMGTOTAL)

total_crop_damage<- crop_damage %>% arrange(-crop_damage$CROPDMGTOTAL)

most_prop_damage_events<-head(total_prop_damage,10)
most_prop_damage_events

par(mfrow = c(1,1), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0), cex = 0.8)
barplot(most_prop_damage_events$PROPDMGTOTAL, names.arg = most_prop_damage_events$EVTYPE,las = 3,col="wheat",main="10 events that damage property",ylab="Property Damage (Billions)")
dev.copy(png, "prop_damage.png", width = 480, height = 480)
dev.off()

most_crop_damage_events<-head(total_crop_damage,10)
most_crop_damage_events

par(mfrow = c(1,1), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0), cex = 0.8)
barplot(most_crop_damage_events$CROPDMGTOTAL, names.arg = most_crop_damage_events$EVTYPE,las = 3,col="grey", main="10 events causing the most injuries",ylab="10 events that damage crops (Billions)")
dev.copy(png, "crop_damage.png", width = 480, height = 480)
dev.off()
```
# Results

### 1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

The analysis and graphs indicate the most harmful type of weather phenomena with respect to death are
*Tornados
*Excessive Heat
*Flash Flood
while in terms of injuries are
*Tornados
*TSTM Wind
*Flood


### 2. Across the United States, which types of events have the greatest economic consequences?

The most harmful events in terms of property damage are
*Flood
*Hurricanes/Typhoons
*Tornados
while in terms of crop damage are
*Drought
*Flood
*River Flood

```{r}
```